PAPER DRAFT

### 2021-11-29 Papers/Polymethylation_scores/05_results.Rmd
comments from erin

---
output:
  html_document: default
  word_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

@@

We calculated several summary measures of DNA methylation, including polymethylation scores, global methylation, and epigenetic clocks. Many of the methylation summary measures were correlated with each other (Supplemental Figure 1). For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001). Our primary polymethylation score  measure, which was calculated using coefficients from a cell-type corrected regression, was very slightly correlated estimated cell-type proportion of immune cells (Pearson $rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).

We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2). The correlation across ages was strongest for the primary polymethylation score (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001). The distribution of epigenetic ages shifted between the age 9 and age 15 visit, as expected (Supplemental Figure 3). Among age 9 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))`  and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`. Among age 15 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))` and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`. The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).

In bivariate analyses and multivariable models, we focused on the polymethylation scores and the single top CpG site from prior research, cg05575921 in the *AHRR* gene,, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.

@@

Mothers who reported smoking during pregnancy had lower income to poverty ratios (`r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)`) than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).  Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=<0.001). Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table 2). At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`). At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`).

@@

The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methlation analysis, and child age  (Figure 3; Supplemental Table 2). At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.  At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. A consistent association was observed when stratifying to the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`). In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though attenuated at age 15 (Figure 3; Supplemental Tables 2).

At age 9, prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting forbase model covariates. Similarly, at age 15 prenatal maternal smoking was associated with `r round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)). The magnitude of the association between prenatal maternal smoking and cg05575921 was reduced in European genetic ancestry samples at age 15 (Figure 3). The direction of the association between prenatal maternal smoking and cg05575921 was reversed in Hispanic genetic ancestry sample (Figure 3).

@@

After additionally adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was very consistent (Supplemental Figure 4). Similarly, after additionally adjusting for other prenatal exposures and postnatal smoke exposure, the direction of the association between prenatal maternal smoking and cg055975921 was consistent, although no longer significant at age 15. Children exposed to postnatal smoke but not prenatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke.  (Supplemental Figure 5). Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).

Results were similar when using linear mixed effect models instead of age-stratified models and adjusting for base model covariates (Supplemental Table 3 & 4).

Results were also similar when using polymethylation scores constructed using alternative regression coefficients (see Methods, Supplemental Figure 7). The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.  Prenatal maternal smoking was also associated with other selected *a priori* CpG sites consistently with associations reported in previous meta-analyses (Supplemental Table 2, Supplemental Figure 8).

@@

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5). We considered classification when using DNA methylation summary measures alone. We also considered classification when using DNA methylation summary measures in addition to a base model of child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two components of genetic ancestry. When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.

At age 9, including the polymethylation score  significantly improved the base model accuracy (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).  At age 15, similarly, including the polymethylation score significantly improved the base model accuracy (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).

At age 9, the base model with the polymethylation score had a larger AUC than the base model with cg05575921 (AUC base model + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`). At age 15, this was also true (AUC for base model + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`). Accuracy was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).

@@

Results were similar when using other coefficients as the weights in construction of the polymethylation scores. However, the AUC was always significantly higher when using the coefficients from a model which incorporated cell-type control as weights (Supplemental Table 6).

>>>>

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

@@

We calculated several summary measures of DNA methylation, including polymethylation scores for smoke exposure, global DNA methylation, and epigenetic clocks. Many of the methylation summary measures were correlated with each other (Supplemental Figure 1). For example, as expected, the percent of estimated immune cells was perfectly inversely correlated with the percent of estimate epithelial cells (Pearson $rho$=`r cor(modeldata$Epithelial.cells_saliva, modeldata$Leukocytes_saliva)`; *P* value<0.0001). Our polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $rho$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value, 3)`).

We compared epigenetic measures between the age 9 and age 15 visit among the `r n_twovisit` individuals with data from both visits (Supplemental Figure 2). The correlation across ages was strongest for the polymethylation score for prenatal smoke exposure (Pearson $\rho$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P* value<0.0001). Global DNA methylation (`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P* value<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: `r round(cor(model_twovisit$pediatric_Age_9, model_twovisit$pediatric_Age_15), 2)`; *P* value<0.0001, GRIM: `r round(cor(model_twovisit$grim_Age_9, model_twovisit$grim_Age_15), 2)`; *P* value<0.0001) The distribution of epigenetic ages shifted higher between the age 9 and age 15 visit, as expected (Supplemental Figure 3). The pediatric clock more closely reflected chronological age than the GRIM clock. Among age 9 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_9, na.rm=T))`  and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_9, na.rm=T))`. Among age 15 samples, the mean estimated age from the pediatric clock was `r round(mean(modeldata_wide$pediatric_Age_15, na.rm=T))` and the mean estimated age from the GRIM clock was `r round(mean(modeldata_wide$grim_Age_15, na.rm=T))`. The distribution of estimated cell-type proportions also shifted between visits while the distribution of global methylation and polymethylation scores were more consistent (Supplemental Figure 3).

In bivariate analyses and multivariable models, we focused on the polymethylation score for prenatal smoke exposure and the single top CpG site from prior research, cg05575921 in the *AHRR* gene, as hypothesized biomarkers, and used global methylation and the pediatric clock as negative controls.

@@

Mothers who reported smoking during pregnancy had lower income to poverty ratios (`r modeldata_people%>%filter(smkPreg_binary=='Yes')%>%summarise(mean(cm1inpov))%>%round(2)`) than those who did not ( `r modeldata_people%>%filter(smkPreg_binary=='No')%>%summarise(mean(cm1inpov))%>%round(2)`).  Mothers who smoked were more likely to report prenatal alcohol use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g2_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%), prenatal drug use (`r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, m1g3_YesNoPreg), margin=1))[1, 2]%>%round(2)*100`%) and postnatal smoking (`r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[2, 2]%>%round(2)*100`% vs `r with(modeldata_people, prop.table(table(smkPreg_binary, PostnatalMaternalSmokingAny), margin=1))[1, 2]%>%round(2)*100`%) than mothers who did not report prenatal maternal smoking (Table 1).  Children of mothers who reported smoking during pregnancy were more likely to be of European genetic ancestry (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[2, 2], 2)*100`%) than children of mothers who did not (`r round(with(modeldata_people, prop.table(table(smkPreg_binary, ancestry), margin=1))[1, 2], 2)*100`%, *P* value=<0.001). Children of mothers who reported smoking during pregnancy had higher prenatal maternal smoking polymethylation scores than children of mothers who did not at both the age 9 (`r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) and age 15 (`r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)` vs `r round(t.test(SSnewbornCT_center_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`, *P* value<0.001) visits (Table 1, Supplemental Table 2). At age 9, children exposed to prenatal maternal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers (`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`%, *P* value=`r round(t.test(cg05575921_Age_9~smkPreg_binary, data=modeldata_wide)$p.value, 2)`). At age 15, children exposed to prenatal smoking had lower DNA methylation at cg05575921 (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group Yes'], 2)`%) than children of non-smoking mothers, although this difference was not significant (`r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$estimate['mean in group No'], 2)`% *P* value= `r round(t.test(cg05575921_Age_15~smkPreg_binary, data=modeldata_wide)$p.value, 2)`). Epigenetic age from the pediatric and GRIM clocks and global DNA methylation did not differ between children exposed vs unexposed to prenatal maternal smoking.

@@

The association between prenatal maternal smoking and the polymethylation score was also observed in multivariable models, adjusting for base model covariates of child sex, maternal income to poverty ratio at baseline, proportion of salivary immune cells, sample plate from DNA methylation analysis, child age and the first two principal components of genetic ancestry  (Figure 3; Supplemental Table 2). At age 9, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score.  At age 15, prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score. A consistent association was observed when stratifying by genetic ancestry. In the African genetic ancestry sample (n= `r modeldata_people%>%filter(ancestry=='African ancestry')%>%nrow()`). In the African genetic ancestry sample at age 9, prenatal maternal smoking was associated with a `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='Sustained smoking polymethylation score (newborns, cell-type controlled)' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)` standard deviation higher prenatal smoke exposure polymethylation score. The direction and magnitude of the association was similar in the European and Hispanic genetic ancestry samples at age 9, though the effect size was attenuated and no longer statistically significant at age 15 (Figure 3; Supplemental Tables 2).

At age 9, prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting for base model covariates. Similarly, at age 15 prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)). Prenatal maternal smoking remained significantly associated with a decrease in cg05575921 DNA methylation in the African genetic ancestry sample at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`) although the association was not significant at age 15  (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`; Figure 3). In the European genetic ancestry sample, the association was consistent in direction at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`). In the Hispanic genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Hispanic ancestry'", 'smkPreg_binaryYes', 2)`) (Figure 3).

@@

After additionally adjusting for other prenatal exposures and postnatal smoke exposure, the association between prenatal maternal smoking and the polymethylation score was robust (Supplemental Figure 4). Similarly, after adjusting for other prenatal exposures and postnatal smoke exposure, the direction of the association between prenatal maternal smoking and cg055975921 was consistent, although no longer significant at age 15. Very few children who were exposed to maternal smoking prenatally were unexposed to postnatal smoke (Supplemental Figure 5). However children exposed only to postnatal smoke did not have higher polymethylation scores than children unexposed to both postnatal and prenatal smoke (Supplemental Figure 5). Results were also similar when controlling for surrogate variables instead of known covariates (Supplemental Figure 6; Supplemental Table 2).

Results from linear mixed effect models were similar to age-stratified and base covariate adjusted linear models (Supplemental Table 3 & 4).

Results were also similar when using polymethylation scores constructed with alternative regression coefficients (see Methods, Supplemental Figure 7). For our main analysis we used coefficients from a regression of sustained smoking exposure and DNA methylation in newborn cord blood with cell-type control. As sensitivity analyses, we used coefficients from regressions of: sustained smoking exposure and DNA methylation in newborn cord blood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children without cell-type control, and any smoking exposure and DNA methylation in newborn cord blood without cell-type control [@Joubert:2016ue]. The association between prenatal maternal smoking and the polymethylation score using regression coefficients from newborn cord blood with cell-type controls was the strongest.

In addition to cg055975921 in the *AHHR* gene, we tested the association of four other *a priori* probe sites identified from previous meta-analyses. We replicated the association of prenatal maternal smoking with these CpG sites (Supplemental Table 2, Supplemental Figure 8).

@@

Next, we compared the accuracy of different DNA methylation summary measures for classifying prenatal maternal smoking using receiver operating curves (Figure 4A; Supplemental Table 5). We estimated classification of prenatal maternal smoking when using DNA methylation summary measures alone. We also estimated prenatal maternal smoking classification when using DNA methylation summary measures in addition to a base model covariates (child sex, maternal income to poverty ratio at baseline, child age at DNA methylation measurement, estimated immune cell proportion, plate from methylation processing and the first two genetic principal components). When used without DNA methylation measures, these base model variables had an area under the curve (AUC) of `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 9')%>%pull(auc)` at age 9 and `r rct1 %>% filter(modeltype=='Base model+No methylation' & childteen=='Age 15')%>%pull(auc)` at age 15.

At age 9, including the polymethylation score  significantly improved the base model covariates classification (AUC: `r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 9')%>%pull(auc)`, *P* value comparing to base model<0.001).  At age 15, similarly, including the polymethylation score significantly improved the base model covariates classification (AUC:`r rct1%>%filter(modeltype=='Base model+Sustained smoking polymethylation score (newborns, cell-type controlled)'&childteen=='Age 15')%>%pull(auc)` , *P* value<0.001).

At age 9, the base model covariates with the polymethylation score had a larger AUC than the base model covariates with cg05575921 (AUC base model covariates + cg05575921: `r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(auc)`; *P* value =`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 9')%>%pull(5)`). At age 15, this was also true (AUC for base model covariates + cg05575921=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(auc)`,  *P* value=`r rct1%>%filter(modeltype=='Base model+AHRR: cg05575921'&childteen=='Age 15')%>%pull(5)`; Supplemental Table 6). Accurate classification was not improved by including polymethylation scores from age 9 and age 15 together (Figure 4B).

@@

Classification of prenatal maternal smoke exposure when using other coefficients as the weights in construction of the polymethylation scores was similar to the results when using coefficients from cell-type controlled regressions of sustained prenatal maternal smoking and DNA methylation in cord blood (Supplemental Table 7).
